---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# shinydo

<!-- badges: start -->
<!-- badges: end -->

The goal of `shinydo` is to guide Shiny craftsmen from "Hello World"
to Shiny software engineering via a practical narrative. It was part of a `R` and
`Shiny` workshop internal to one of my customers.

## Installation

You can install the development version of `shinydo` from [GitHub](https://github.com/) with:

``` {r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("RudyPastel/shinydo")
```

## How to use shinyDemo

Obviously, you should try the app out.

```{r example, eval = FALSE}
shinydo::showPresentation()
```

More importantly however, read, analyse and evaluate the code and its quality.

Comments, suggestions and customer inquiries are welcome. Contact [Dr Rudy Pastel](mailto:rudy.pastelgmail.com).
