---
title: "shinydo"
author: Dr Rudy Pastel
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{shinydo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, 
  echo = TRUE,
  eval = FALSE,
  comment = "#>"
)
```

`shinydo` is the path away from the pains of haphazard scripts toward shiny 
engineering. `shinydo` shows many doors but the craftsman has to go through them.

```{r, echo=FALSE, eval=TRUE, layout="l-page", fig.height=2,fig.cap='The path'}
DiagrammeR::grViz(diagram = "images/graphScripts/path.gv", width = '100%')
```

Come along as we journey the fabled Google trend index project.

## Key ideas to remember

1. R is a functional language: use functions and shiny modules
1. R/Shiny development is software engineering: use version control and packages. 
1. Especially in group, think processes through: use the Golem framework.

A curriculum referencing study material concludes this presentation


## R, in the off chance you do not know

R is a first choice programming language / environment for:

- data analysis via statistics, machine learning, data mining ...
- visualisation of tables, maps, networks ...
- publication of reports, book, reproducible research ...

Shiny brings the analysis results to users via a gui.


## Create "Google trend index" on your own

The boss tasks you with developing Google trend indices. You proudly submit 
the following shiny app.

```{r,echo=TRUE,eval=FALSE}
shinydo::runInternalApp(appName = 'googleTrendIndex_V0', display.mode = 'showcase')
```

Because the app is tiny, the code seems acceptable.

* The code is divided in sections with clear purposes
* The code breathes i.e. spans many short lines so as to be easy to read
* Comments are plentiful and informative
* Variable names are descriptive

None the less, its structure will  not survive the extension required by initial 
success. 

### Google, Bing, and Yahoo

Assume the boss is happy with the Google trend index app and now wants you 
to include the other big search engines i.e. Bing and Yahoo. The
trend indices should be displayed independently for a more flexible analysis. How
do you do it?

```{r,echo=TRUE,eval=FALSE}
shinydo::pauseAndThinkAboutIt(seconds = 30)
message('Share your ideas via the common chat.')
```

Unfortunately, an all to frequent answer is the quick and dirty copy-paste-rename
method.

```{r,echo=TRUE,eval=FALSE}
shinydo::runInternalApp(appName = 'googleTrendIndex_V1', display.mode = 'showcase')
```

The quick and dirty copy-paste-rename method typically results in the following ills.

* The length and structurelessness of the code beget repulsion and tedium.
* Repetition killed readability and maintainability.
* The intertwining of business logic with shiny logic cause confusion.

Do not do that! Once the code reaches this sorry state, no one wants to touch it.
Usually, the code is then quietly abandoned and dies off silently. But not always...

### Google, Bing, Yahoo, Baidu and Yandex then top 10

The boss is happy with the top 3 and now wants a top 10 analysis. You wisely
negotiate to first provide a top 5 as an opportunity to restructure the code and
enable future extensions. How do you do it?

```{r,echo=TRUE,eval=FALSE}
shinydo::pauseAndThinkAboutIt(seconds = 30)
message('Share your ideas via the common chat.')
```

This time, you decide to leverage the functional nature of R and the possibility
to split the app into two files and a folder of support functions.

* Define the front end of your app in ui.R 
* Define the back end of your app in server.R
* Define the support functions of your app in the R/ folder
  * Implement the business logic in pure R functions
  * Implement the shiny logic as shiny modules

In preparation to this new approach, let us first expend on Shiny modules.

##### Shiny modules

Shiny modules avoid the duplication of shiny code by virtue of being R functions
encapsulating Shiny logic. Practically, a shiny module is a pair of functions. 
One defines a ui building block and the one defines its server logic. Because
they can be tested and nested, modules allow the decomposition of the app is manageable
pieces. The function `shiny::shinyAppTemplate()` provides an exemple.

```{r,echo=TRUE,eval=FALSE}
shinydo::runInternalApp(appName = 'shinyModuleDemo', display.mode = 'showcase')
```

The example module is implemented as the pair of functions `exampleModuleUI` and
`exampleModuleServer`. The module i.e. the pair of functions is used twice: once
with identifier `examplemodule1` and once with identifier `examplemodule2`. The
module defines the logic exactly once and reuses it every time. Its implementation
makes it obvious.

```{r,eval=TRUE,echo=FALSE,results='asis'}
moduleCode = readLines(con = file.path(
  shinydo::getInternalAppFolder(), 'shinyModuleDemo', 'R', 'example-module.R')
  )
moduleCode = c('```{r,echo=TRUE,eval=FALSE}', moduleCode, '```')
cat(paste(moduleCode, collapse = '\r\n'))
```


### Google, Bing, Yahoo, Baidu and Yandex

The top 5 app structure is so crystal clear, there is no need to split its code
between ui.R and server.R

```{r,echo=TRUE,eval=FALSE}
shinydo::runInternalApp(appName = 'googleTrendIndex_V2', display.mode = 'showcase')
```

The plot function is pure R and can be developed, tested and reused independently
from the app.

```{r,eval=TRUE,echo=FALSE,results='asis'}
moduleCode = readLines(con = file.path(
  shinydo::getInternalAppFolder(), 'googleTrendIndex_V2', 'R', 'plotTrendIndex.R')
  )
moduleCode = c('```{r,echo=TRUE,eval=FALSE}', moduleCode, '```')
cat(paste(moduleCode, collapse = '\r\n'))
```

The module code is as simple as the very first app. Simplifying it with submodules
is possible but would be an overkill.

```{r,eval=TRUE,echo=FALSE,results='asis'}
moduleCode = readLines(con = file.path(
  shinydo::getInternalAppFolder(), 'googleTrendIndex_V2', 'R', 'indexTrendModule.R')
  )
moduleCode = c('```{r,echo=TRUE,eval=FALSE}', moduleCode, '```')
cat(paste(moduleCode, collapse = '\r\n'))
```

### Top 10 search engine trend index

Extending the app to a top 10 is now trivial.

```{r,echo=TRUE,eval=FALSE}
shinydo::runInternalApp(appName = 'googleTrendIndex_V3', display.mode = 'showcase')
```

The boss is happy and praises you. You bask in glory.

## Share "Google trend index" with your users

Your work is very successful and popular: multiple colleagues from multiple teams
want to use your results and even build on them for their own analysis. The boss
wants you to make it happen and make the team / them proud. How do you do it?

```{r,echo=TRUE,eval=FALSE}
shinydo::pauseAndThinkAboutIt(seconds = 30)
message('Share your ideas via the common chat.')
```

Please, do not:

* Share code via mail or a share folder
* Assume users will understand your code and just make it work
* Organise "quick" introduction meetings to explain the code and install the 
code on your users' computers
* Have even "quicker" meetings to fix bugs on their machines

Instead, do this:

* Remember that future you is your first user and be kind
* Convert your code into a R package 
  * Actually, R packages should be the default code structure. 
* Document and test all functions
* Select the functions accessible to users and demonstrate them in a tutorial 
* Put your package in an accessible code repository such as [github](https://github.com/).
* Share a code snippet downloading the package and opening the tutorial.
* Improve the package based on feedback and put new versions in the repo

For your users, accessing your code should be as simple as those lines.

```{r}
# Get started
install.packages('shinydo')
vignette(package = 'shinydo', 'shinydo')

# Run the app
shinydo::startTrendIndexDashboard ()

# Query documentation
?shinydo::getTrendIndexData
?shinydo::plotTrendIndex

# Use key functions
trend_data = shinydo::getTrendIndexData(searchEngine = 'Google')
trend_data = subset(x = trend_data, subset = type == 'A')
shinydo::plotTrendIndex(date = trend_data$date, 
                        close = trend_data$close, 
                        smoother = TRUE, 
                        smootherSpan = 0.3)
```

### R package 101

The `devtools` package is your friend and creates the package skeleton for you.

```{r,eval=TRUE,echo=TRUE}
packageFolder = file.path(tempdir(), 'demoPackage')
devtools::create(path = packageFolder, open = FALSE)
```


Just write the right information in the right place.

```{r,eval=TRUE,echo=FALSE}
fs::dir_tree(path = packageFolder, recurse = FALSE)
```
It is simple:

1. In the `DESCRIPTION` file, list the packages required to run your code.
1. In `R`, put your documented functions, marking those accessible to users.
1. Build the documentation file with `devtools::document()`.
1. Build the package file with `devtools::build()`.
1. You are done.

The final `DESCRIPTION` may look like this.

```{r,eval=TRUE,echo=FALSE,results='asis'}
moduleCode = readLines(con = './../DESCRIPTION')
moduleCode = c('```{r,echo=TRUE,eval=FALSE,cap="DESCRIPTION"}', moduleCode, '```')
cat(paste(moduleCode, collapse = '\r\n'))
```

A documented function file in the `R` folder may look like this.

```{r,eval=TRUE,echo=FALSE,results='asis'}
moduleCode = readLines(con = './../R/plotTrendIndex.R')
moduleCode = c('```{r,echo=TRUE,eval=FALSE,cap="plotTrendIndex.R"}', moduleCode, '```')
cat(paste(moduleCode, collapse = '\r\n'))
```


```{r,eval=TRUE,echo=FALSE,results='asis'}
rdPath = list.files(path = './../man/', pattern = 'plotTrendIndex', full.names = TRUE)
htmlPath = tempfile(pattern = 'plotTrendIndex.html')
tools::Rd2HTML(Rd = rdPath, out = htmlPath, package = "shinydo", stylesheet = "R.css")
htmlLines = readLines(con = htmlPath)
cat(paste(htmlLines[-(1:3)], collapse = '\r\n'))
```

### Shiny package 101

This is a R package with Shiny modules and a function to start the app.


```{r,eval=TRUE,echo=FALSE,results='asis'}
moduleCode = readLines(con = './../R/appFunctions.R')
moduleCode = c('```{r,echo=TRUE,eval=FALSE,cap="plotTrendIndex.R"}', moduleCode, '```')
cat(paste(moduleCode, collapse = '\r\n'))
```

### Main benefits of R packages

* The package is a well established and documented standard
  * Developers can rely on robust tools, processes and conventions to save time
  * Users know how to read the documentation, use the functions and build on them
* The package enables many quality improving features
  * Documentation of functions and tutorial for usages
  * Test of functions, including shiny modules
  * Reproducible research / data analysis

## Develop "Google trend index" as a team

Your works enjoys laudatory acclaim! A group of power users gathers and calls for
more. The boss gives you carte blanche to extend the analysis and publish the tool
internally. How do you set up your team? How do you organise development?


```{r,echo=TRUE,eval=FALSE}
shinydo::pauseAndThinkAboutIt(seconds = 30)
message('Share your ideas via the common chat.')
```

Shiny is great for proofs of concepts. Shiny can deliver software product too!
This requires a mindset shift from data analysis and visualisation to software engineering.

### Know your limits and hire expertise

Transforming a Shiny app into a software product requires new skills.
Know your limits and hire the relevant expertise in:

* Architecture
* Cyber-security
* Dev Ops
* User experience / User interface

The IT department can introduce you to the correct people for those tasks. Dedicated
R tools exists and should be used. You may not have the culture to think about them.
The experts will prompt you to look for them.

### Rely on thought through processes

Software emerges from the interactions of people, tools, code and processes.
The [Golem](https://engineering-shiny.org/index.html) framework is all about the
process of building an application that will later be sent to production. 

#### Design

1. Sketch the user interface with users
1. Style the user interface with users
1. Draw the architecture 

#### Prototype


1. Prototype the Ui with a place holding back-end for users to test and approve
1. Prototype the business logic to understand the user's need and document it

#### Build

1. Organise your code to keep business logic, shiny logic and style apart
1. Define processes w.r.t. to version control, development and tests
1. Build the Ui and gather user approval
1. Implement, document and test the business logic function
1. Combine business and shiny logic into modules

#### Strengthen

1. Test the business logic functions
1. Test the modules
1. Test load resistance
1. Test in a reproducible environment
1. Test automatically thanks to continuous integration

#### Deploy

1. Decide how your users will access your work:
    1. Download and install package from repository
    1. Access running app via a secure url
1. Design, implement and automate the delivery process to:
    1. [github](https://github.com/)
    1. Your department private server
    1. ...
    
#### Test 101

To test a function, do not manually call it with random input. Instead, write a 
set of well chosen test scenarios to be automatically and always run. As usual,
a dedicated package helps: `usethis`.

```{r,echo=TRUE,eval=TRUE,warning=FALSE}
setwd(packageFolder)
usethis::use_r(name = "myNewFunction")    # creates and opens R/myNewFunction.R
usethis::use_test(name = "myNewFunction") # creates and opens tests/testthat/test-myNewFunction.R
```

The core of unit testing is making sure that for a given input, the function 
behaves as you expect.

```{r}
test_that("multiplication works", {
  expect_equal(2 * 2, 4)
})

test_that("myNewFunction works", {
  expect_equal(myNewFunction(wellChosenInput), expectedOutput)
  expect_error(myNewFunction(wrongInput), 'Error message')
})

```


First, implement the new function and the test in parallel around an easy set of
inputs. Then add tests of increasing difficulties for edge cases, missing input,
input of the wrong kind... `tests/testthat.R` runs all the tests.

```{r,eval=TRUE,echo=FALSE}
fs::dir_tree(path = packageFolder, recurse = TRUE)
```



## R curriculum

```{r, echo=FALSE,eval=TRUE, results='asis'}
res <- knitr::knit_child("./r-curriculum.Rmd", quiet = TRUE)
res = gsub(pattern = '##', replacement = '###', x = res)
cat(res, sep = '\n')
```

```{r,eval=FALSE, child="./r-curriculum.Rmd"}

```

